Delayed effects of climate on vital rates lead to demographic divergence in Amazonian forest fragments

### 2021-08-11 doc/paper.Rmd
updated descriptive statistics related text. #72

```{r include=FALSE}
surv_by_year_habitat <-
  model_data %>% 
  group_by(year, habitat) %>% 
  summarize(prop_surv = mean(surv)) %>% 
    arrange(prop_surv)
surv_by_year_habitat

surv_by_year <-
  model_data %>% 
  group_by(year) %>% 
  summarize(prop_surv = mean(surv)) %>% 
    arrange(prop_surv)
surv_by_year
```

@@

*Survival:* Across all plots, the proportion of plants surviving was lowest in the 2003--2004 transition year ($P_{surv} = 0.92$).
This coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d) and was preceded by a drop in average plant size in the 2002--2003 transition year (i.e. negative growth in Figure \@ref(fig:eda)a).
The lowest survival for 1-ha fragment plots ($P_{surv} = 0.93$) was for the 2005--2006 transition year, which encompassed a moderate drought in October 2005 and and wetter than average conditions (SPEI \> 0.5) in December 2005 and January 2006 (Figure \@ref(fig:eda)b,d).
The lowest survival for continuous forest was in 2004 ($P_{surv}=0.91$).
When summarizing across years, plots, and plant sizes, the survival probability of *Heliconia acuminata* was similarly high in both continuous forest and fragments ($P_{surv}=0.95$; Figure \@ref(fig:eda)b; the overlapping 85% CI of model intercepts indicate no significant difference).
However, survival in both habitats was size dependent ($p < 0.001$ for the effect of log-transformed plant size in year *t* on survival in year *t+1* in both habitats).
The survival probability of large plants approached 1 in both habitat types (Figure \@ref(fig:size-plot)b), but the smallest plants had higher survival in 1-ha fragments.

@@

*Size:* Plants in continuous forest had an average of 2.9 shoots (± 1.8 SD) and were on average 40.6 cm tall (± 26.5 SD).
Plants in 1-ha fragments had on average `r format((1 - (2.5/2.9))*100, digits = 3)`% fewer shoots (2.5 ± 1.5 SD) and were `r format((1 - (36.2/40.6))*100, digits = 3)`% shorter (36.3 cm ± 24.1 SD).
Because our proxy for plant size was the product of these two metrics, plants in continuous forest were on average 34% larger than those in forest fragments (150 ± 175 SD vs. 112 ± 141 SD, respectively), with fragments having proportionately fewer large plants (Figure \@ref(fig:size-plot)d).
This difference was not significant, however (overlap in 84% CI of model intercepts), and the disparity in plant size---which was most pronounced in the initial years of our surveys---diminished over time (Figure \@ref(fig:eda)a).

@@

*Flowering:* The overall proportion of plants that flowered was very low.
While it was 27% higher in continuous forest than 1-ha fragments (`r summary_stats$flwr_mean[2]` ± `r summary_stats$flwr_sd[2]` vs. `r summary_stats$flwr_mean[1]` ± `r summary_stats$flwr_sd[1]`, respectively), this difference was not statistically significant (84% CIs of model intercepts overlapped).
The observed disparity was largely due to the fact that flowering is also size-dependent ($p < 0.001$ in both habitats), with the probability of flowering increasing dramatically once plants reached the threshold size of about 148 (i.e., log(size) \> 5 in Figure \@ref(fig:size-plot)c).
Despite the flowering probability of large plants being greater in fragments than continuous forest, populations in fragments had proportionately fewer plants above the reproductive size threshold (Figure \@ref(fig:size-plot)d).
The most striking difference between habitat types coincided with a severe drought in 2003, when the percentage of flowering reproductive-sized plants was 28% in continuous forest vs. only 13.6% in 1-ha fragments (Figure \@ref(fig:eda)c).

>>>>

```{r include=FALSE}
surv_by_year_habitat <-
  model_data %>% 
  group_by(year, habitat) %>% 
  summarize(prop_surv = mean(surv)) %>% 
    arrange(habitat, prop_surv)
surv_by_year_habitat

surv_by_year <-
  model_data %>% 
  group_by(year) %>% 
  summarize(prop_surv = mean(surv)) %>% 
    arrange(prop_surv)
surv_by_year

model_data %>% summarize(mean(surv))
```

@@

*Survival:* Across all plots, the proportion of plants surviving was lowest in the 2003--2004 transition year ($P_{surv} = 0.93$ in both continuous forest and forest fragments).
This coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d) and was preceded by a drop in average plant size in the 2002--2003 transition year (i.e. negative growth in Figure \@ref(fig:eda)a).
When summarizing across years, plots, and plant sizes, the survival probability of *Heliconia acuminata* was similarly high in both continuous forest and fragments ($P_{surv}=0.95$; Figure \@ref(fig:eda)b; the overlapping 85% CI of model intercepts indicate no significant difference).
However, survival in both habitats was size dependent ($p < 0.001$ for the effect of log-transformed plant size in year *t* on survival in year *t+1* in both habitats).
The survival probability of large plants approached 1 in both habitat types (Figure \@ref(fig:size-plot)b), but the smallest plants had higher survival in 1-ha fragments.

@@

```{r include=FALSE}
summary_stats
```

*Size:* Plants in continuous forest had an average of 2.9 shoots (± 1.8 SD) and were on average 39.8 cm tall (± 26.2 SD).
Plants in 1-ha fragments had on average `r format((1 - (2.45/2.85))*100, digits = 3)`% fewer shoots (2.5 ± 1.4 SD) and were `r format((1 - (35.57/39.75))*100, digits = 3)`% shorter (35.6 cm ± 23.8 SD).
Because our proxy for plant size was the product of these two metrics, plants in continuous forest were on average 32% larger than those in forest fragments (144 ± 176 SD vs. 109 ± 139 SD, respectively), with fragments having proportionately fewer large plants (Figure \@ref(fig:size-plot)d).
This difference was not significant, however (overlap in 84% CI of model intercepts), and the disparity in plant size---which was most pronounced in the initial years of our surveys---diminished over time (Figure \@ref(fig:eda)a).

@@

```{r include=FALSE}
summary_stats
```

*Flowering:* The overall proportion of plants that flowered was very low.
While it was 31% higher in continuous forest than 1-ha fragments (`r summary_stats$flwr_mean[2]` ± `r summary_stats$flwr_sd[2]` vs. `r summary_stats$flwr_mean[1]` ± `r summary_stats$flwr_sd[1]`, respectively), this difference was not statistically significant (84% CIs of model intercepts overlapped).
The observed disparity was largely due to the fact that flowering is also size-dependent ($p < 0.001$ in both habitats), with the probability of flowering increasing dramatically once plants reached the threshold size of about 148 (i.e., log(size) \> 5 in Figure \@ref(fig:size-plot)c).
Despite the flowering probability of large plants being greater in fragments than continuous forest, populations in fragments had proportionately fewer plants above the reproductive size threshold (Figure \@ref(fig:size-plot)d).
The most striking difference between habitat types coincided with a severe drought in 2003, when the percentage of flowering reproductive-sized plants was 28% in continuous forest vs. 17% in 1-ha fragments (Figure \@ref(fig:eda)c).

```{r include=FALSE}
repro_size <- model_data %>% filter(flwr == 1) %>% pull(size) %>% quantile(.1, na.rm = TRUE)

model_data %>%
  filter(size >= repro_size) %>% 
  group_by(habitat, year) %>%
  summarize(prop_flwr = sum(flwr, na.rm = TRUE)/n()) %>% 
    arrange(year) %>% 
  pivot_wider(names_from = habitat, values_from = prop_flwr) %>% 
  mutate(diff = `1-ha` - CF) %>% 
  arrange(desc(abs(diff)))
```
